---
title: "Class 6 Testing Graduations (Solution)"
author: "Yubo Rasmussen"
format: docx
editor: visual
---

## Introduction

This notebook will start with a section on standard table graduation, followed by testing graduation and goodness of fits.

## Standard Table Graduation `Standard.r`

This section illustrates standard table graduation using the **Gompertz model with Poisson errors** and compares different modeling approaches applied to **CMI data**. First, we load the required dataset, which contains mortality data for different years and age groups.

```{r}
# URL for CMI_read.r on github
CMI_Deaths_url <- "https://raw.githubusercontent.com/yubae-bit/F79SU/main/CMI%20and%20HMD%20data%20sets/CMI_Deaths.csv"
CMI_Exposures_url <- "https://raw.githubusercontent.com/yubae-bit/F79SU/main/CMI%20and%20HMD%20data%20sets/CMI_Exposures.csv"
CMI_Read_url <- "https://raw.githubusercontent.com/yubae-bit/F79SU/main/CMI%20and%20HMD%20data%20sets/CMI_read.r"

# Download the necessary files
download.file(CMI_Deaths_url, destfile = "CMI_Deaths.csv", mode = "wb")
download.file(CMI_Exposures_url, destfile = "CMI_Exposures.csv", mode = "wb")

# Now source the R script
source("CMI_read.r")
```

------------------------------------------------------------------------

### Select Data for Standard Table Graduation

We begin by selecting mortality data for the year 2000, focusing on ages 40 to 80. This data will be used to fit our standard table graduation model.

```{r}

```

### Fit Gompertz Model with Poisson Errors

Now, we fit a Poisson regression model to estimate mortality rates using the Gompertz function. The offset for exposure ensures proper modeling of mortality rates per unit exposure.

```{r}

```

### Visualization of Standard Table Graduation

The fitted model is plotted to visualize the estimated mortality rates against observed values. This helps assess how well the model fits the data.

```{r}
```

### Save Standard Table Graduation on Log Scale

We store the fitted log mortality rates for further comparison with other datasets.

```{r}
```

### Comparison with Fresh Data (Year 2005)

Next, we retrieve mortality data for the year 2005 and compare it against the standard table from 2000 to observe potential mortality trends.

```{r}
```

### Graduation of 2005 Data using Gompertz Model

To understand how mortality rates evolve, we fit a new Gompertz model for the 2005 dataset and compare it with the standard table graduation.

```{r}
```

### Fit GLM of DTH on Standard Table

Now, we fit a generalized linear model (GLM) to estimate the relationship between deaths and the previously fitted standard table log mortality rates.

```{r}
```

### Weighted Least Squares Graduation

In addition to the Poisson regression approach, we perform a weighted least squares regression to account for variability in exposure.

```{r}
```

### Comparison of Standard and Data Graduation

We compare different graduation approaches through multiple visualizations to assess the effectiveness of different modeling techniques.

```{r}
```

#### Interpretation of the plots

-   **Panels A & C** show that both **Gompertz and GLM models fit the mortality data well**.
-   **Panel B** confirms that **the standard table graduation closely follows observed data**.
-   **Panel D** highlights **differences between GLM and weighted least squares approaches**, which may suggest potential refinements in modeling mortality trends.

### Conclusion

-   The Gompertz Poisson model is effective for mortality graduation.
-   Comparison of 2000 and 2005 data allows assessment of changes in mortality trends.
-   Weighted least squares provides an alternative fit that accounts for exposure size.

## Functions for Testing Graduation `Test_GoF.r`

### Chi-squared

```{r}
#     Chi-square test
#
#     Arguments: Z = residuals Z.x
#                Npar = number of fitted parameters
#
```

### Standardized deviations test

#### Equal Area Test

```{r}
#     Equal area test
#
#     Arguments: Z = residuals Z.x
#                N = Number of cells
#

```

#### Equal Width Test

```{r}
#     Equal width test
#
#     Arguments: Z = residuals Z.x
#                W = Width of internal cells
#
```

### Sign Test

```{r}
#     Sign test
#
#     Argument: Z = residuals Z.x
#
```

### Change of Sign Test

```{r}
#     Change of sign test
#
#     Argument: Z = residuals Z.x
#
```

### Runs Test

```{r}
#     Runs test
#
#     Arguments: Z = residuals Z.x
#
#     This function calls the subsiduary function Runs( )
#

```

#### Using a Permutation Test

```{r}
#     Runs test using a permutation test
#
#     Arguments: Z = residuals Z.x
#                n = number of samples - defaults to 1000 if n is not set
#     Calls the function Runs.test( )
#

```

### Serial Correlation test

```{r}
#     Serial correlation test
#
#     Arguments: Z = residuals Z.x
#
```

## Example from Chapter 6 `Chap_6_Eg.r`

```{r}
#
#    Example in Chapter 6 to illustrate statistical tests
#
#    Input data, draw plots, fit model and calculate Z.x
#

#
#    Chis^2 test
#
#
#    Standardised deviations test
#
#    Equal width cells
#

#    Equal area cells
#
#
#    Sign test
#
#
#    Change of sign test
#
#
#    Runs test
#
#
#    and in its permutation form
#
#
#    Serial correlation test
#
#  

```

How would you interpret each test?

## Aside: Using simulations in `ChiTest_Sim.r`

```{r}
#
#     Chis2 tests or standardized deviations test via simulation
#
#     Z contains the residuals
#     N is the number of cells
#
#
#

```

How would you interpret this?
