---
title: "Class 7: Lee Carter"
author: "Yubo Rasmussen"
format: docx
editor: visual
---

```{r setup, include=FALSE}
# Load required library
library("gnm")
```

## **Lab Overview**

In this lab, we will explore the **Lee-Carter mortality model**, a widely used method for modeling and forecasting mortality rates.

## **1. Load Data and Define Variables**

```{r}

```

## **2. Fit the Lee-Carter Model**

```{r}

```

**💡 Question 2:** Why do we use Poisson regression in this model? What does `offset=log(Exp.V)` achieve?

## **3. Adjust for Identifiability Constraints**

```{r}

```

## **4. Compute Fitted Mortality Rates**

```{r}

```

## **5. Visualizing the Model Parameters**

```{r}

```

## **6. Forecasting Future Mortality Rates**

We now forecast mortality rates using a **random walk with drift model** for (\kappa\_t).

```{r}

```

## **7. Accounting for Parameter Uncertainty**

We also calculate **parameter uncertainty** in our forecasts:

```{r}

```
