---
title: "Class 2: Kaplan-Meier"
author: "Yubo Rasmussen"
format: html
editor: visual
---

# Introduction

This notebook demonstrates how to compute and visualize Kaplan-Meier estimates of the survival function using the `survival` package in R. Each example explores a different dataset and scenario to highlight the flexibility of Kaplan-Meier survival analysis.

## Example 1: Time to Discontinuation of a Medical Device

The data represent an experiment studying the time to discontinuation of a medical device. Observations include both complete and censored data, indicated by `Censor`.

### Load the Data
```{r}

# Time to discontinuation (in weeks)

# Censoring indicator (1 = event occurred, 0 = censored)

# Create a survival object
```

### Kaplan-Meier Estimate
```{r}
# Fit the Kaplan-Meier survival curve

# Summary of the survival object
```

### Plot the Survival Curve
```{r}
```

---

## Example 2: Acute Myelogenous Leukaemia (AML) Data

This example uses AML data where:
- `1` indicates a relapse occurred.
- `0` indicates the patient left the study or was still in remission.

### Load and Explore the Data
```{r}
# Load the aml dataset

# Create a survival object
```

### Kaplan-Meier Estimate by Group
```{r}
# Fit the Kaplan-Meier survival curve by group (x)

# Summary of the survival object
```

### Plot the Survival Curve
```{r}

# Add legend
```

---

## Example 3: Rat Drug Trial

This example involves survival times of rats in a drug trial. Observations include both events and censored data.

### Load the Data
```{r}
# Time to event (in days)

# Censoring indicator (1 = event occurred, 0 = censored)

# Ensure lengths match

# Create a survival object
```

### Kaplan-Meier Estimate
```{r}
# Fit the Kaplan-Meier survival curve

# Summary of the survival object
```

### Plot the Survival Curve
```{r}
```