---
title: "Class 2: Kaplan-Meier"
author: "Yubo Rasmussen"
format: docx
editor: visual
---

# Introduction

This notebook demonstrates how to compute and visualize Kaplan-Meier estimates of the survival function using the `survival` package in R. Each example explores a different dataset and scenario to highlight the flexibility of Kaplan-Meier survival analysis.

## Example 1: Time to Discontinuation of a Medical Device

The data represent an experiment studying the time to discontinuation of a medical device. Observations include both complete and censored data, indicated by `Censor`.

### Load the Data

```{r}
library(survival)

# Time to discontinuation (in weeks)

# Censoring indicator (1 = event occurred, 0 = censored)

# Create a survival object
```

**Questions and Answers**:

1\. **What is the purpose of the `Surv()` function?**
2\. **Why do we need the censoring indicator?** 
### Kaplan-Meier Estimate

```{r}
# Fit the Kaplan-Meier survival curve

# Summary of the survival object
```

**Questions and Answers**:

1\. **What does `~ 1` mean in this context?** 

2\. **What does `conf.type = "plain"` mean in this context?** 

3\. **What does the Kaplan-Meier estimate represent?** 

### Plot the Survival Curve

```{r}
```

**Questions and Answers**:

1\. **Why is it important to visualize the Kaplan-Meier curve?** 

2\. **What do the confidence intervals indicate on the plot?** 

------------------------------------------------------------------------

## Example 2: Acute Myelogenous Leukaemia (AML) Data

This example uses AML data where: - `1` indicates a relapse occurred. - `0` indicates the patient left the study or was still in remission.

### Load and Explore the Data

```{r}
# Load the aml dataset

# Create a survival object
```

**Questions and Answers**:

1\. **What does `status` represent in the AML dataset?** 

2\. **How does the Kaplan-Meier estimate handle censored patients in this dataset?** 

### Kaplan-Meier Estimate by Group

```{r}
# Fit the Kaplan-Meier survival curve by group (x)

# Summary of the survival object
```

**Questions and Answers**:

1\. **What does stratifying by `x` achieve in this analysis?** 
2\. **What key differences might you expect between groups in this example?** 

### Plot the Survival Curve

```{r}

# Add legend
```

**Questions and Answers**:

1\. **What does the legend add to this plot?**

2\. **Why are different line types used for groups?** 

------------------------------------------------------------------------

## Example 3: Rat Drug Trial

This example involves survival times of rats in a drug trial. Observations include both events and censored data.

### Load the Data

```{r}
# Time to event (in days)

# Censoring indicator (1 = event occurred, 0 = censored)

# Ensure lengths match

# Create a survival object
```

**Questions and Answers**:

1\. **Why is it important to check that the lengths of `Time` and `Censor` match?** 

2\. **What assumptions are made about censored data in this example?** 

### Kaplan-Meier Estimate

```{r}
# Fit the Kaplan-Meier survival curve

# Summary of the survival object
```

### Plot the Survival Curve

```{r}

```

**Questions and Answers**:

1\. **What features of the survival curve might indicate the effect of the drug?** 