---
title: "Tutorial 5: Graduation I"
author: "Yubo Rasmussen"
format: docx
editor: visual
---

## Tutorial 5: Graduation I

### Question 1

Extract the data for females ages 40 to 80 for the year 2005 from the England and Wales (E&W) data set. Use R to fit Gompertz type models; i.e., for $μ_x$ assume that $\log(μ_x)$ is linear in age $x$ while for $q_x$ assume that $\text{logit}(q_x)$ is linear in $x$.

#### Load the Data

```{r}
# Define URLs for all required data files
EngWales_Deaths_url <- "https://raw.githubusercontent.com/yubae-bit/F79SU/main/CMI%20and%20HMD%20data%20sets/EngWales_Deaths.txt"
EngWales_Exposures_url <- "https://raw.githubusercontent.com/yubae-bit/F79SU/main/CMI%20and%20HMD%20data%20sets/EngWales_Exposures.txt"
EngWales_Read_url <- "https://raw.githubusercontent.com/yubae-bit/F79SU/main/CMI%20and%20HMD%20data%20sets/EngWales_read.r"

# Download the necessary files
download.file(EngWales_Deaths_url, destfile = "EngWales_Deaths.txt", mode = "wb")
download.file(EngWales_Exposures_url, destfile = "EngWales_Exposures.txt", mode = "wb")
download.file(EngWales_Read_url, destfile = "EngWales_read.r", mode = "wb")

# Now source the R script
source("EngWales_read.r")

```

#### Extract the Data

```{r}
#
#    Extract female data, ages 40 to 80, year 2005.
#

#
#   It will be useful to have this as a function so
#
Main.plot <- function( ) plot(AGE, Obs, xlab = "Age", ylab = "log(mortality)",
     main = "Gompertz law: England & Wales, female, ages 40 to 80, year 2005",
     pch = 16, col = "blue")
```

#### Part (a)

The model for $μ_x$ using weighted least squares to estimate the parameters.

```{r}
#
#  (a) Weighted normal for mu_x
#

```

#### Part (b)

The model for $μ_x$ using a GLM with Poisson errors to estimate the parameters.

```{r}
#
#  (b) GLM with Poisson errors
#

```

#### Part (c)

The model for $q_x$ using weighted least squares to estimate the parameters. (Hint: You need $\text{Var}[\text{logit}(\hat{q}_x)] ≈ 1/d_x$).

```{r}
#
#   (c) Weighted normal for q_x
#

```

#### Part (d)

The model for $q_x$ using a GLM with binomial errors to estimate the parameters.

```{r}
#
#  (d) GLM with binomial errors
#

```

#### Part (e)

Compare the GLM graduations of $μ_x$ and $q_x$.

```{r}
#
#  (e) 
#
```

#### Part (f)

Obtain the standardised residuals, $z_x$, in each of the four graduations and plot the $z_x$ against $x$. Comment informally on what these plots tell you.

```{r}
#
#  (f)
#
#  Normal residuals
#


```

### Question 2

#### Part (a)

Extract the male data for age 65 for all years from the E&W data set. Plot the maximum likelihood estimates of $\log(μ_t)$ against year, $t$. Consider Gompertz style graduations of $μ_t$ with

$$
\log(\mu_t) = a_0 + a_1t + a_2t^2 + ... + q_pt^p 
$$

for $p = 1, 2, 3, 4.$ Use the `glm( )` function and Poisson errors to estimate the regression coefficients. Compare the various fits with appropriate plots. *Hint: You will do well to use powers of* $t − \bar{t}$ *in the calls of `glm( )`; see section 5.1.4.*

```{r}
#
#    Extract female data, age 65, years 1841 to 2005 (all years).
#

#
#   It will be useful to have this as a function so
#
Main.plot <- function( ) plot(Year, Obs, ylab = "log(mortality)", cex = 0.5,
     main = "England & Wales, male, age 65", pch = 16, col = "blue")

#
#   Define some powers of year
#

```

#### Part (b)

Repeat (a) for age 20.

```{r}
#
#   Age 20 can be done with a change of data and the same program.
#   The labelling needs altered (although this too can be made automatic).
#

```

#### Part (c)

Note for interest the amazing changes in mortality that have taken place from 1841 to 2005.
